Towards Practical First-Order Model Counting

### 2024-08-02 doc/paper/paper.tex
Preparing for a new round of experiments
Some of the preparations for a new set of experiments (including scaling FastWFOMC domains exponentially, since it's faster now)
Everything related to the new set of experiments (except for the results, which are too big to be stored here)
The full experimental results are too big to fit into a repo

\section{Experimental Evaluation}\label{sec:experiments}
% TODO: this section is apparently "hard to follow"

% TODO: Compilation time is included in the experimental results, and the
% differences between compilation and inference time are delineated in the
% second-to-last paragraph of Section 6.

% TODO: It would not have been feasible to increase domain sizes by 1 all the
% way up to 300000. Thus, the experimental results are presented in two parts,
% where the first part examines scalability with small domain sizes, and the
% second part tests the limits of each algorithm within a given timeout.

% TODO: (here or in the introduction) C++ code generation is necessary for any
% runtime experiments because without it the algorithm is "incomplete" as
% explained above.

% TODO: The asymptotic analysis of the total number of arithmetic operations
% required to compute the FOMC of these formulas was already performed in
% previous work.

% TODO: Since FastWFOMC does not support many-sorted logic, our experiments were
% restricted to formulas with a single domain. However, many of the counting
% problems examined in [1] become equivalent (in the sense that they generate
% the same integer sequences) when restricted to a single domain. Moreover, it
% is challenging to compare Crane2 and FastWFOMC on a larger set of problems
% because each problem has to be represented in two (fairly different) logics:
% \FO{} and \UFO{}.

We compared \Cranetwo{}\footnote{\textsc{Crane} has two modes that decide how
  compilation rules are applied to formulas: one that uses greedy search and
  another that combines greedy and breadth-first search. In our experiments, we
  refer to these modes as \Cranegreedy{} and \Cranebfs{}, respectively.} with
\textsc{FastWFOMC} and \textsc{ForcLift} on two problems previously considered
by \citeauthor{DBLP:conf/kr/DilkasB23}~\shortcite{DBLP:conf/kr/DilkasB23}. The
first problem is the function-counting problem from \cref{example:functions},
and the second problem is the bijection-counting problem described below. It is
important to note that comparing \Cranetwo{} and \textsc{FastWFOMC} on a more
substantial set of benchmarks is challenging because there is no automated way
to translate a formula in \FO{} or \Ctwo{} into \UFO{}, or even check if such an
encoding is possible. We ran the experiments on an AMD~Ryzen~7~5800H processor
with \SI{16}{\gibi\byte} of memory and Arch Linux~6.8.2-arch2-1 operating
system. \textsc{FastWFOMC} was executed using Python~3.8.19 with
Python-FLINT~0.5.0.

@@

\begin{figure}[t]
  \centering
  \input{plot}
  \caption{The running time of WFOMC algorithms. Note that the $y$-axis is on a
    logarithmic scale.}\label{fig:plot}
\end{figure}
% TODO: turn this into two separate plots for clarity. Maybe also attach labels
% to their corresponding lines

% TODO: no overfull boxes (might be the fault of the figure)

% TODO: must not use 'input' command. instead, compile to a pdf

In the first experiment, we ran each of the four algorithms once on each
combination of benchmark and domain size, ranging from one up to and including
35. \Cref{fig:plot} shows the total runtime values as a measure of the overall
performance of each algorithm. We also separately tracked compilation and
inference times and commented on them where relevant. It is worth noting that
\Cranebfs{} can handle more instances than either \textsc{ForcLift} or
\Cranegreedy{}, including the bijection-counting problem in our experiments as
well as similar formulas examined previously~\cite{DBLP:conf/kr/DilkasB23}.

As shown in \cref{fig:plot}, the runtimes of all compilation-based algorithms
remained practically constant, in contrast to the rapidly increasing runtimes of
\textsc{FastWFOMC}. Although the search/compilation part is slower in
\Cranetwo{} than in \textsc{ForcLift}, the difference is negligible. The
runtimes of the FOKC algorithms appear constant because, for these counting
problems and domain sizes, compilation time dominates inference time (remember
that compilation time is independent of domain sizes). Indeed, the maximum
inference time of \Cranebfs{} and \Cranegreedy{} across these experiments is
only \SI{4}{\milli\second}. The runtimes of \Cranetwo{} have lower variation
than those of \textsc{ForcLift} because we compile the formula anew for each
domain size with \textsc{ForcLift}, whereas with \Cranetwo{}, we compile it once
and reuse the resulting C++ program for all domain sizes.
% TODO: The lines for Crane2 in Figure 4 appear to be completely flat whereas
% the line for ForcLift has some slight fluctuations. We refer to these
% fluctuations as variance.

\begin{table}[t]
  \centering
  \begin{tabular}{lrr}
    \toprule
    Algorithm & Bijections & Functions \\
    \midrule
    \Cranebfs{} & \num{e4} & \num{3e5} \\
    \Cranegreedy{} & --- & \num{3e5} \\
    \textsc{FastWFOMC} & 28 & 32 \\
    \textsc{ForcLift} & --- & 143 \\
    \bottomrule
  \end{tabular}
  \caption{The maximum domain sizes that each algorithm can handle within
    \SI{45}{\second}. A dash symbol `---' indicates that the algorithm was
    unable to find a solution. The domain sizes for \Cranetwo{} are accurate up
    to the first digit, while the domain sizes for the other domain sizes are
    exact.}\label{table:results}
\end{table}

For the second experiment, we examined the maximum domain size the algorithms
can handle in at most \SI{45}{\second} of total runtime. As shown in
\cref{table:results}, \Cranebfs{} can scale to domain sizes \num{357} and
\num{9375} times larger than \textsc{FastWFOMC} for the bijection-counting and
function-counting problems, respectively. It is important to note that while
\textsc{ForcLift} may have runtime performance comparable to \Cranetwo{}, its
finite-precision arithmetic cannot represent model counts for domain sizes
greater than 143 and returns $\infty$ instead.

>>>>

\section{Experimental Evaluation}\label{sec:experiments}

% TODO: 1. Introduction
% Objective: Briefly state the goals of the experiments.
% Overview: Summarize the structure of the section.

\begin{itemize}
  \item Note that it would not make sense to experimentally evaluate \Cranetwo{}
        without the C++ code generation described in \cref{sec:cpp} because the
        algorithm would only produce mathematical definitions of functions.
\end{itemize}

% TODO: 2. Experimental Setup

% Environment: Describe the hardware and software environment (e.g.,
% specifications of the computer, operating system, libraries, and tools used).

\begin{itemize}
  \item The experiments were run on an Intel Skylake \SI{2.4}{\giga\hertz} CPU
        with \SI{188}{\gibi\byte} of memory and CentOS~7 operating system. C++
        programs were compiled using the Intel C++ Compiler 2020u4.
        \textsc{FastWFOMC} was run on Julia~1.10.4, while all other algorithms
        were run on the Java Virtual Machine 1.8.0\_201.
\end{itemize}

% Datasets: Provide details about the datasets used in the experiments,
% including size, source, and any preprocessing steps.

\begin{itemize}
  \item The asymptotic analysis of the total number of arithmetic operations
        required to compute the FOMC of some of these formulas was already
        performed in previous work~\cite{DBLP:conf/kr/DilkasB23}.
  \item Since FastWFOMC does not support many-sorted logic, our experiments were
        restricted to formulas with a single domain. However, many of the
        counting problems examined in [1] become equivalent (in the sense that
        they generate the same integer sequences) when restricted to a single
        domain. Moreover, it is challenging to compare Crane2 and FastWFOMC on a
        larger set of problems because each problem has to be represented in two
        (fairly different) logics: \FO{} and \UFO{}.
  \item It is important to note that comparing \Cranetwo{} and
        \textsc{FastWFOMC} on a more substantial set of benchmarks is
        challenging because there is no automated way to translate a formula in
        \FO{} or \Ctwo{} into \UFO{}, or even check if such an encoding is
        possible.
\end{itemize}

% Baselines: List and describe the baseline methods or systems used for
% comparison.

\begin{itemize}
  \item \textsc{Crane} has two modes that decide how compilation rules are
        applied to formulas: one that uses greedy search and another that
        combines greedy and breadth-first search. In our experiments, we refer
        to these modes as \Cranegreedy{} and \Cranebfs{}, respectively.
\end{itemize}

% Metrics: Define the performance metrics used to evaluate the results (e.g.,
% accuracy, precision, recall, F1-score, runtime).

\begin{itemize}
  \item Compilation time is included in the experimental results, and the
        differences between compilation and inference time are delineated in the
        second-to-last paragraph of Section 6.
  \item We run each algorithm on each benchmark on domains of size
        $2^{0}, 2^{1}, 2^{2},\dots$ until the algorithm times out after
        \SI{1}{\hour}, runs out of memory, or (in the case of \textsc{ForcLift})
        returns $\infty$.
\end{itemize}

% TODO: 3. Results and Analysis
% Presentation of Results: Use tables, graphs, and charts to present the experimental results clearly.
% Comparison with Baselines: Compare your method’s results with the baselines.
% Discussion: Analyze and interpret the results, highlighting the strengths and potential weaknesses of your approach.
% Insights: Provide insights or patterns observed from the results.
% Challenges: Discuss any challenges or unexpected outcomes.
% Acknowledgement: Honestly acknowledge any limitations of your experiments or method.

\begin{itemize}
  \item It is important to note that while \textsc{ForcLift} may have runtime
        performance comparable to \Cranetwo{}, its finite-precision arithmetic
        cannot represent model counts for domain sizes greater than 143 and
        returns $\infty$ instead.
\end{itemize}

% TODO: 8. Conclusion
% Summary: Briefly summarize the key findings from the experimental results.
% Future Work: Suggest directions for future research based on your findings.

We compare \Cranetwo{} with \textsc{FastWFOMC} and \textsc{ForcLift} on two
problems previously considered by
\citeauthor{DBLP:conf/kr/DilkasB23}~\shortcite{DBLP:conf/kr/DilkasB23}. The
first problem is the function-counting problem from \cref{example:functions},
and the second problem is the bijection-counting problem described below.

@@

\begin{figure*}[t]
  \centering
  \includegraphics{plot.pdf}
  \caption{The runtime of the algorithms as a function of the domain size. Note
    that both axes are on a logarithmic scale.}\label{fig:plot}
\end{figure*}

\Cref{fig:plot} shows the total runtime values as a measure of the overall
performance of each algorithm. We also separately tracked compilation and
inference times and commented on them where relevant. It is worth noting that
\Cranebfs{} can handle more instances than either \textsc{ForcLift} or
\Cranegreedy{}, including the bijection-counting problem in our experiments as
well as similar formulas examined previously~\cite{DBLP:conf/kr/DilkasB23}.

As shown in \cref{fig:plot}, the runtimes of all compilation-based algorithms
remained practically constant, in contrast to the rapidly increasing runtimes of
\textsc{FastWFOMC}. Although the search/compilation part is slower in
\Cranetwo{} than in \textsc{ForcLift}, the difference is negligible. The
runtimes of the FOKC algorithms appear constant because, for these counting
problems and domain sizes, compilation time dominates inference time (remember
that compilation time is independent of domain sizes). Indeed, the maximum
inference time of \Cranebfs{} and \Cranegreedy{} across these experiments is
only \SI{4}{\milli\second}. The runtimes of \Cranetwo{} have lower variation
than those of \textsc{ForcLift} because we compile the formula anew for each
domain size with \textsc{ForcLift}, whereas with \Cranetwo{}, we compile it once
and reuse the resulting C++ program for all domain sizes.
